Numerical simulation and entropy dissipative cure of the carbuncle instability for the shallow water circular hydraulic jump

### 2020-11-12 numerics_paper/version2/numerical_chj.tex
update

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[hmargin=0.9in, vmargin=1.25in]{geometry}
\usepackage[babel,german=quotes]{csquotes}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{caption}
\usepackage{amsmath}          % writing mathematical formulas
\usepackage{amsthm}           % writing mathematical theorems 
\usepackage{amssymb}          % writing mathematical symbols
\usepackage{bm, amsfonts}     % writing bold mathematical symbols
\usepackage{xcolor}
\usepackage{fixmath}
\usepackage{tikz}
\usepackage{bm}
\usepackage{makecell}
\usepackage{mathdots}
\usepackage{algpseudocode}
\usepackage{algorithm}

\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}

@@

\subsection{Our contribution}
\begin{itemize}
\item We present a new problem where the cc instability occurs. In our opinion, this is a more challenging 
  problem than the standard bow problem since we expect the solution to be non-steady. 
  We aim to preserve instabilities that are inherently present in the CHJ for certain regimes. And, 
  at the same time, we aim to prevent the cc instability. 
\item We explore the use of entropy dissipation and stability as cure of the cc instability for the SWEs. 
\item We propose specific methods to blend the highly dissipative and robust Rusanov's RP and the 
  accurate 3-wave Roe's solver for the SWEs. In particular we propose methods that induce dissipation of entropy
  near the shocks and other methods that guarantee entropy stability. 
\item We demonstrate that these schemes prevent the formation of cc instabilities not only 
  in the well known bow shock problem, but also in the CHJ while preserving other important features of 
  the solution. 
\end{itemize}

@@

\section{Numerical methods}
We consider the implementation of the Lax-Wendroff method in Clawpack. 
For a one-dimensional problem, the method can be written as follows:
\begin{align*}
  U_i^{n+1}=U_i^n-\frac{\Delta t}{\Delta x}\left[\A\Delta Q_{i-1/2}]
\end{align*}
where XX. In this section we focus on the first order method; i.e., we consider

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[hmargin=0.9in, vmargin=1.25in]{geometry}
\usepackage[babel]{csquotes}
\usepackage{subfig}
\usepackage{graphicx}
\usepackage[colorlinks=true,citecolor=blue,urlcolor=blue]{hyperref}
\usepackage{caption}
\usepackage{amsmath}          % writing mathematical formulas
\usepackage{amsthm}           % writing mathematical theorems 
\usepackage{amssymb}          % writing mathematical symbols
\usepackage{bm, amsfonts}     % writing bold mathematical symbols
\usepackage{xcolor}
\usepackage{fixmath}
\usepackage{tikz}
\usepackage{bm}
\usepackage{makecell}
\usepackage{mathdots}
\usepackage{algpseudocode}
\usepackage{algorithm}

\newcommand{\W}{{\mathcal W}}
\newcommand{\A}{{\mathcal A}}
\newcommand{\bff}{{\bf f}}
\newcommand{\bfr}{{\bf r}}
\newcommand{\bfF}{{\bf F}}
\newcommand{\bfu}{{\bf u}}
\newcommand{\bfv}{{\bf v}}
\newcommand{\bfq}{{\bf q}}
\newcommand{\bfx}{{\bf x}}
\newcommand{\sgn}{{\rm sgn}}
\newcommand{\Rus}{{\rm Rus}}
\newcommand{\Roe}{{\rm Roe}}

%\newdefinition{rmk}{Remark}
\newtheorem{remark}{Remark}

@@

\subsection{Our contribution}
\begin{itemize}
\item We present a new problem where the cc instability occurs. In our opinion, this is a more challenging 
  problem than the standard bow problem since we expect the solution to be non-steady. 
  We aim to preserve instabilities that are inherently present in the CHJ for certain regimes. And, 
  at the same time, we aim to prevent the cc instability. 
\item We explore the use of entropy dissipation and stability as cure of the cc instability for the SWEs. 
\item We propose specific methods to blend the highly dissipative and robust Rusanov's RP and the 
  accurate 3-wave Roe's solver for the SWEs. In particular we propose methods that induce dissipation of entropy
  near the shocks and other methods that guarantee entropy stability. 
\item We demonstrate that these schemes prevent the formation of cc instabilities not only 
  in the well-known bow shock problem, but also in the CHJ while preserving other important features of 
  the solution. 
\end{itemize}

@@

\clearpage
\section{Numerical methods}
In this work, we consider Riemann-based finite volume methods.
Considering a one dimensional hyperbolic conservation law,
the cell average of the numerical solution given by a 
finite volume scheme (with forward Euler) can be written as follows:
\begin{align}\label{first-order_FV}
  \bfu_i^{n+1}=\bfu_i^n-\frac{\Delta t}{\Delta x}\left[\bfF_{i+1/2}-\bfF_{i-1/2}\right],
\end{align}
where $\bfF_{i+1/2}=\bfF_{i+1/2}(u^n)$ is a numerical flux; i.e., $\bfF_{i+1/2}$ is 
a discretization of the flux $\bff(\bfu(x_{i+1/2},t^n))$. 
This numerical flux is composed of two parts, a consistent term that approximates 
the physical flux at the interface between cells $i$ and $i+1$, and a stabilization term. 
For instance, the well-known Rusanov's numerical flux is given by 
\begin{align}\label{rusanov_num_flux}
  \bfF_{i+1/2}(\bfu)=\frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2} - \lambda_{i+1/2}(\bfu_{i+1}-\bfu_i),
\end{align}
where $\lambda_{i+1/2}$ is an upper bound on the wave speed associated with the Riemann problem 
composed of the states $\bfu_i$ and $\bfu_{i+1}$. 
%If $z=u$ in \eqref{rusanov_num_flux}
%; i.e., if the Rusanov's numerical flux is obtained considering 
%the cell averages of the solution 
%and if the time step is small enough, 
Method \eqref{first-order_FV} with Rusanov's flux \eqref{rusanov_num_flux}
is positivity preserving and entropy stable, see for instance 
\cite{perthame1996positivity,tadmor2003entropy,guermond2016invariant}
However, the accuracy of the numerical flux, and hence of the method, is 
limited to first-order. 
Two alternatives are common to improve the accuracy properties of Rusanov's method.
The numerical flux $\bfF_{i+1/2}$ can be approximated based on high-order polynomial 
reconstructions (or combinations of them) around the interface $i+1/2$. 
For smooth solutions, this strategy leads to arbitrarily high-order
methods. Alternatively, second-order accuracy can be obtained by removing some of the 
numerical dissipation introduced by the stabilization component of the numerical flux. 
The second-order Lax-Wendroff method \cite{lax1959systems} is a popular example of this approach. 
In this work we consider an adaptation of Lax-Wendroff method,
sometimes known as Lax-Wendroff-LeVeque \cite{leveque1997wave,leveque2002finite}, 
that incorporates limiters into the second-order correction. 
These high-order corrections to Rusanov's method require careful treatment to guarantee 
the solution is still positivity preserving and/or entropy stable.

The accuracy of a given first- or second-order method depends heavily on the Riemann solver. 
In particular, the Rusanov's Riemann solver is highly dissipative. In contrast, the well-known 
Roe's Riemann solver produces more accurate solutions. 
%
As we discussed in the introduction, finite volume methods with 
Roe's Riemann solvers are prone to the carbuncle instability for the CHJ. 
In contrast, if we use the Rusanov's 
Riemann solver, no carbuncles appear; however, many other important and physical features of 
the solution are dissipated, unless the grid is highly refined. 
Our main objective is to study the CHJ and to propose a Riemann solver
that is carbuncle free but do not dissipate important features of the solution. 
To do that we propose a Riemann solver that blends Rusanov's and Roe's solvers. 
Our blending criteria is based on the residual of the entropy (in)equality. 
Similar ideas have been explored before to solve the Euler equations XX 
and the shallow water equations XX. In these references, the authors use different 
blending criteria.

\subsection{Lax-Wendroff-LeVeque method}
Let us now review the second-order correction to method \eqref{first-order_FV} implemented in 
Clawpack \cite{clawpack} and PyClaw \cite{pyclaw-sisc}.
We refer to \cite{leveque1997wave, leveque2002finite} for a detailed description of the method. 
We consider only a one-dimensional problem; in the 
numerical experiments in \S\ref{sec:num} we use second-order Strang's splitting \cite{strang1968construction} 
to extend the method to multiple dimensions.

The finite volume method \eqref{first-order_FV} can be written in terms of right- 
and left-going waves emanating from the interfaces $i-1/2$ and $i+1/2$, respectively. 
The net contribution of these waves is called right- and left-going fluctuations. 
Upon defining $z^+:=\frac{1}{2}(z+|z|)$ and $z^-:=\frac{1}{2}(z-|z|)$, 
the fluctuations are given by
\begin{align}\label{fluct}
  %\A^+\Delta Q_{i-1/2} := \frac{1}{2}\sum_p\left(s_{i-1/2}^p+|s_{i-1/2}^p|\right)\W_{i-1/2}^p, \quad
  %\A^-\Delta Q_{i+1/2} := \frac{1}{2}\sum_p\left(s_{i+1/2}^p-|s_{i+1/2}^p|\right)\W_{i+1/2}^p,
  \A^+\Delta Q_{i-1/2} := \sum_p\left(s_{i-1/2}^p\right)^+\W_{i-1/2}^p, \qquad
  \A^-\Delta Q_{i+1/2} := \sum_p\left(s_{i+1/2}^p\right)^-\W_{i+1/2}^p,
\end{align}
where $\W_{i\pm 1/2}^p$ and $s_{i\pm 1/2}^2$ are the waves and their corresponding speeds, respectively. 
The waves and their speed of propagation (to compute the fluctuations) are tightly related to Riemann solvers. 
In \S\ref{sec:riemann_solvers} we review the definition of the waves and speeds associated with 
Rusanov's and Roe's Riemann solvers.
Based on fluctuations, method \eqref{first-order_FV} can be written as 
\begin{align}\label{first-order_via_fluct}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}\left[\A^+\Delta Q_{i-1/2}+\A^-\Delta Q_{i+1/2}\right].
\end{align}

The second-order Lax-Wendroff-LeVeque method removes some of the numerical dissipation in 
\eqref{first-order_via_fluct} via a flux correction at each interface. The method is written as follows:
\begin{align}\label{second-order_via_fluct}
  u_i^{n+1} = u_i^n-\frac{\Delta t}{\Delta x}\left[\A^+\Delta Q_{i-1/2}+\A^-\Delta Q_{i+1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde{F}_{i+1/2}-\tilde{F}_{i-1/2}\right],
\end{align}
where 
\begin{align*}
  \tilde{F}_{i\pm 1/2}=\frac{1}{2}\sum_p|s_{i\pm 1/2}^p|\left(1-\frac{\Delta t}{\Delta x}|s_{i\pm 1/2}^p|\right)\tilde\W_{i\pm 1/2}^p
\end{align*}
are the flux corrections at the interfaces $i\pm 1/2$. Here $\tilde{\W}_{i\pm 1/2}^p$ is a 
TVD-limited version of $\W_{i\pm 1/2}^p$. 
In this work, we consider the min-mod limiters, we refer to \cite{leveque1997wave} for more details. 
If no limiters are used 
($\tilde \W_{i\pm 1/2}^p\equiv \W_{i\pm 1/2}^p$), \eqref{second-order_via_fluct} becomes
the standard second-order Lax-Wendroff scheme, which in terms of fluctuation-like quantities 
is given by
\begin{align*}
  u_i^{n+1}=u_i^n-\frac{\Delta t}{\Delta x}
\left[
  \frac{1}{2}\left(s_{i-1/2}^p+\frac{\Delta t}{\Delta x}|s_{i-1/2}^p|^2\right)\W_{i-1/2}^p
  +
  \frac{1}{2}\left(s_{i+1/2}^p-\frac{\Delta t}{\Delta x}|s_{i+1/2}^p|^2\right)\W_{i+1/2}^p
\right].
\end{align*}
To implement method \eqref{second-order_via_fluct}, 
we need to choose a Riemann solver. Although the scheme is second-order 
regardless of the Riemann solver, the accuracy and robustness depends upon the 
Riemann solver. The approximate Riemann solver based on Rusanov's flux is robust and known to be carbuncle free. 
However, it adds large amounts of numerical diffusion that, for relative coarse grids, 
might dissipate other physical and important features of the solution. If the apprroximate Riemann solver 
is based on the Roe's average, the amount of numerical diffusion is highly reduced; however, 
such an approach is prone to develop the carbuncle instability for certain problems. 
Many carbuncle fixes to the Roe's solver have been proposed for the Euler equations. 
In the context of the shallow water equations, we review in \S\ref{sec:riemann_solvers} 
the Riemann solver proposed in \cite{kemm2014note}.

\section{Riemann solvers}\label{sec:riemann_solvers}

For the shallow water equations in two-dimensions, the structure of the exact (and some 
approximated) Riemann solvers contains three waves. This is the case, for example, of the well-known 
Roe's Riemann solver, see for instance \cite{roe1981approximate} and 
\cite[\S 15.3.3 and \S 21.7]{leveque2002finite}. 
Simpler approaches are also possible. 
For instance, methods \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}
with the Rusanov's numerical flux are based on only two waves that 
propagate with the same speed in opposite directions.
We refer to \cite{ketcheson2020riemann} and references therein for 
a full and detailed review of these two Riemann solvers.

\subsection{Roe's average}\label{sec:roe}
The Roe's Riemann solver is based on evaluating the flux Jacobian using the Roe's average
\begin{align}\label{roe_average}
  \bar h_{i-1/2}=\frac{1}{2}(h_{i-1}+h_i), \qquad 
  \hat u_{i-1/2}=\frac{\sqrt{h_{i-1}}u_{i-1}+\sqrt{h_i}u_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}, \qquad
  \hat v_{i-1/2}=\frac{\sqrt{h_{i-1}}v_{i-1}+\sqrt{h_i}v_i}{\sqrt{h_{i-1}}+\sqrt{h_i}}.
\end{align}
Let 
\begin{align*}
\hat A_{i-1/2} =
\begin{bmatrix}
  0 & 1 & 0 \\
  -\hat u_{i-1/2}^2+g\bar h_{i-1/2} & 2\hat u_{i-1/2} & 0 \\
  -\hat u_{i-1/2} \hat v_{i-1/2} & \hat v_{i-1/2} & \hat u_{i-1/2}
\end{bmatrix}
\end{align*}
be the flux Jacobian evaluated using the Roe's average at the interface $i-1/2$.
The average \eqref{roe_average} is designed to guarantee: 
i)  consistency of the approximate flux Jacobian 
($\hat A_{i-1/2}\rightarrow \bff^\prime(\bfu(x_{i-1/2}))$ as $\bfu_i, \bfu_{i-1}\rightarrow \bfu(x_{i-1/2})$); 
ii) hyperbolicity ($\hat A_{i-1/2}$ is diagonalizable with real eigenvalues);
and iii) conservation ($\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$). 
Let
\begin{align*}
  \hat \bfr^1_{i-1/2} = 
  \begin{bmatrix}
    1 \\ 
    \hat u_{i-1/2}-\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix},
  \qquad 
  \hat \bfr^2_{i-1/2} = 
  \begin{bmatrix}
    0\\ 
    0\\ 
    1
  \end{bmatrix},
  \qquad
  \hat \bfr^3_{i-1/2} = 
  \begin{bmatrix}
    1 \\ 
    \hat u_{i-1/2}+\sqrt{g\bar h_{i-1/2}}\\
    \hat v_{i-1/2}
  \end{bmatrix}
\end{align*}
be the eigenvectors of $\hat A_{i-1/2}$.
The waves for the two-dimensional Riemann solver based on the Roe's average are 
$\W_{i-1/2}^p=\alpha_{i-1/2}^p\hat r_{i-1/2}^p$, with $p=1,2,3$. Here $\alpha_{i-1/2}^p$ is given by
\begin{align}\label{system_for_alphas}
  \left[\hat \bfr^1_{i-1/2} ~\hat \bfr^2_{i-1/2} ~\hat \bfr^3_{i-1/2}\right]
  \begin{bmatrix}
    \alpha^1_{i-1/2} \\
    \alpha^2_{i-1/2} \\
    \alpha^3_{i-1/2}
  \end{bmatrix}
  =\Delta Q_{i-1/2}:=\bfu_i-\bfu_{i-1}.
\end{align} 
%where $\alpha_{i-1/2}^p$ is the $p$-th entry from the solution of 
%\end{align*}
The speed of the waves is given by the eigenvalues of $\hat A_{i-1/2}$
\begin{align*}
  \hat\lambda_{i-1/2}^1=\hat u_{i-1/2}-g\sqrt{g\bar h_{i-1/2}}, \qquad
  \hat\lambda_{i-1/2}^2=\hat u_{i-1/2}, \qquad
  \hat\lambda_{i-1/2}^3=\hat u_{i-1/2}+g\sqrt{g\bar h_{i-1/2}}. \\
\end{align*}

The waves and speeds at interface $i+1/2$ are computed similarly. 
The fluctuations are given by \eqref{fluct} with $s_{i\pm 1/2}^p=\hat\lambda_{i\pm 1/2}^p$ 
Finally, the first- and second-order methods based on Roe's Riemann solver are given by 
\eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively. 
We close this review by remarking that the conservation property 
$\hat A_{i-1/2}(\bfu_i-\bfu_{i-1})=\bff(\bfu_i)-\bff(\bfu_{i-1})$ implies that 
\begin{align}\label{cons_roe}
  \A^-\Delta Q_{i-1/2}+\A^+\Delta Q_{i-1/2}=\sum_p \hat\lambda_{i-1/2}^p\W_{i-1/2}^p=\bff(\bfu_{i})-\bff(\bfu_{i-1}),
\end{align}
a property we will use in the next section.

\subsection{Rusanov's Riemann solver}\label{sec:rusanov}
The Rusanov's Riemann solver considers only two waves that emanate from each interface, 
say for instance interface $i-1/2$, and travel with the same speed $\lambda_{i-1/2}^{\max}\geq 0$ 
but in opposite directions. 
The speed of propagation $\lambda_{i-1/2}^{\max}$ is an upper bound on the maximum wave speed
associated with the corresponding Riemann problem between the states $\bfu_{i-1}$ and $\bfu_{i}$. 
In this work, we define $\lambda_{i-1/2}^{\max}$ via the guaranteed upper bound from \cite[Prop. 3.7]{azerad2017well}. 
The waves are given by 
\begin{align}\label{rusanov_waves}
  \W_{i-1/2}^{1,\Rus}=\bfu_{i-1/2}-\bfu_{i-1}, \qquad 
  \W_{i-1/2}^{2,\Rus}=\bfu_{i}-\bfu_{i-1/2}, 
\end{align}
where $\bfu_{i-1/2}$ is a middle state in the solution of the Riemann problem. The middle state is obtained
by imposing
\begin{align}\label{rus_cons}
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{1,\Rus}+\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus} = \bff(\bfu_{i})-\bff(\bfu_{i-1}),
\end{align}
which guarantees the solver is conservative. 
The fluctuations are given by \eqref{fluct} with $s_{i-1/2}^1=-\lambda_{i-1/2}^{\max}$
and $s_{i-1/2}^2=\lambda_{i-1/2}^{\max}$, and the first- and second-order methods based 
on Rusanov's solver are given by \eqref{first-order_via_fluct} and \eqref{second-order_via_fluct}, respectively.

We close this section by remarking that the fluctuations associated with Rusanov's solver 
can be expressed in terms of waves and speeds from the Roe's solver. To see this, first consider \eqref{rus_cons}
and rewrite the right-going fluctuation as follows:
\begin{align}\label{rus_rs_aux}
  \A^{+,\Rus}\Delta Q_{i-1/2}:=\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \bff(\bfu_{i})-\bff(\bfu_{i-1}) + \lambda_{i-1/2}^{\max}\left(\W_{i-1/2}^{1,\Rus}+\W_{i-1/2}^{2,\Rus}\right)
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus}.
  %\A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
\end{align}
From \eqref{system_for_alphas} and \eqref{rusanov_waves} we get 
$\W^{1,\Rus}_{i-1/2}+\W^{2,\Rus}_{i-1/2}=\bfu_i-\bfu_{i-1}=\sum_p\W_{i-1/2}^p$. 
Using \eqref{cons_roe}, \eqref{rus_rs_aux} becomes 
\begin{align*}
  \lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\max}
  = \sum_p \left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p
  -\lambda_{i-1/2}^{\max}\W_{i-1/2}^{2,\Rus},
\end{align*}
which implies that 
\begin{subequations}\label{fluct_rus}
\begin{align}
  \A^{+,\Rus}\Delta Q_{i-1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i-1/2}^p + \lambda_{i-1/2}^{\max}\right)\W_{i-1/2}^p,
\end{align}
and similarly, 
\begin{align}
  \A^{-,\Rus}\Delta Q_{i+1/2}=\frac{1}{2}\sum_p\left(\hat \lambda_{i+1/2}^p - \lambda_{i+1/2}^{\max}\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

\subsubsection{HLLEMCC solver}
The HLLEMCC approximate Riemann solver for the shallow water equations was proposed
in \cite{kemm2014note} specifically to deal with carbuncles.  The idea is to adjust
the amount of dissipation applied to the shear wave, adding dissipation in regions
where unphysical carbuncles would appear.  The approximate solution in this
case consists of four waves; the faster waves use the HLLE wave speeds and carry
jumps only in the depth and normal momentum, while the slower waves carry jumps
only in the transverse momentum.  By adjusting the value of $\phi$, this solver
can behave like the HLLE solver (if $\phi=1$) or like the Roe solver (if $\phi=0$).

The middle states $q^1_m, q^2_m$ are the same as those used for the HLLE solver.
Requiring conservation of the transverse momentum gives
\begin{align*}
  q^3_m = \frac{f^3_l - f^3_r - (\hat{v}-\phi\hat{c})q^3_l + (\hat{v}+\phi\hat{c})q^3_r}{2\phi\hat{c}}.
\end{align*}
Here $\hat{v}, \hat{c}$ are the Roe-average normal velocity and sound speed, respectively, whereas the function $\phi$ is given by
\begin{equation}\label{if}
\phi(\theta)=\min\left\lbrace 1,\, \left(\varepsilon\,\theta\,\max\left\lbrace0,\,1-Fr_{u}^{\alpha} \right\rbrace \right)   \right\rbrace^{\beta}
\end{equation}
where $Fr_{u}$ is the Froude number perpendicular to the cell face and $\varepsilon$, $\alpha$, and $\beta$ are parameters that have to be chosen appropriately. For the shallow water equations, authors in \cite{kemm2014note} propose to use $\varepsilon=10^{-3}$ and $\alpha=\beta=\frac{1}{3}$. The value of $\theta$ is calculated as the 2-norm of the residual in the Rankine-Hugoniot condition for the shear wave.

The HLLEMCC solver follows the ideas of the Harten-Hyman entropy fix by splitting the shear wave with speed $\hat{u}$, into two waves moving with speeds $\hat{u}-\phi\hat{c}$ and $\hat{u}+\phi\hat{c}$ (see figure \ref{HLLEMCC}). Its implementation, in terms of the fluctuations, is performed by using the values $\hat{u}^{\pm}$ calculated with an absolute-value function defined as

\begin{align*}
\varphi(\lambda)=\frac{\vert \lambda-\phi(\theta)\hat{c}\,\vert+\vert \lambda+\phi(\theta)\hat{c}\,\vert}{2}
\end{align*}

\begin{figure}
    \center
    \includegraphics[width=0.6\textwidth]{figures/hllemcc.pdf}
    \caption{Structure of the HLLEMCC approximate Riemann solution, with four waves.}
    \label{HLLEMCC}
\end{figure}

\section{Entropy dissipative Riemann solver}\label{sec:ent_diss_rs}
Consider a hyperbolic system of conservation laws 
\begin{align}\label{cons_law}
  \frac{\partial \bfu(\bfx)}{\partial t} + \nabla\cdot \bff(\bfu(\bfx))=0, \quad \bfx\in\Omega,
\end{align}
with a en entropy pair $\left(\eta(\bfu), \bfq(\bfu)\right)$. 
A weak solution of \eqref{cons_law} is called an entropy solution if 
\begin{align}\label{ent_inequality}
  \frac{\partial \eta(\bfu(\bfx))}{\partial t} + \nabla\cdot \bfq(\bfu(\bfx))\leq 0, \quad \bfx\in\Omega,
\end{align}
holds, in a distribution sense, for any entropy pair.
It is desirable for a numerical method to satisfy a discrete version of the entropy inequality 
\eqref{ent_inequality}. This is, however, not true for many standard numerical discretizations. 
In general, an entropy stable scheme introduces nonlinear stabilization, which is 
particularly important for solutions with strong discontinuities that might converge to non-physical 
weak solutions. 
In this work, we propose an entropy stable approximate Riemann solver for the one dimensional 
shallow water equations. We use this Riemann solver as building block within the Lax-Wendroff-LeVeque 
method, as described in \cite{leveque1997wave,leveque2002finite} and implemented in Clawpack.

\subsection{Entropy residual}
Let $S_i$ be the set of cells containing cell $i$ and those that share a face with it. 
Also let $\bfv(\bfu)=\eta^\prime(\bfu)$ be the entropy variable. 
Based on \cite{guermond2011entropy}, we consider 
\begin{align}\label{ent_residual}
  \int_{S_i} \bfv(\bfu) \cdot \left[ \frac{\partial \bfu}{\partial t} + \nabla\cdot \bff(\bfu)\right]d\bfx
  =\int_{S_i} \left[\frac{\partial \eta(\bfu)}{\partial t} + \bfv(\bfu) \cdot \nabla\cdot \bff(\bfu)\right] d\bfx
\end{align}
as a measurement of the entropy production in $S_i$. 
To avoid approximating the time derivative of the entropy, we follow 
\cite{guermond2018second, guermond2018well} and use 
$\int_{K_i} \frac{\partial\eta(\bfu)}{\partial t}d\bfx=-\int_{K_i} \nabla\cdot\bfq(\bfu) d\bfx$, 
which holds in smooth regions. Then we define 
\begin{align}\label{Ri}
  R_i := \frac{N_i}{D_i}, 
\end{align}
with
\begin{align*}
  N_i=
  \left|\int_{S_i} \left[\bfv(\bfu) \cdot \nabla\cdot \bff(\bfu) - \nabla\cdot\bfq(\bfu) \right] d\bfx\right|,
\end{align*}
and $D_i$ being an upper bound of $N_i$ that acts as normalization so that $0\leq R_i\leq 1$.
Note that $N_i\approx 0$ if $\bfu$ is smooth in $S_i$. 
In our implementation, we use
\begin{align*}
  N_i\approx \left|\bfv(\bfu_i)\cdot \int_{S_i}\nabla\cdot \bff(\bfu)d\bfx 
  -\int_{S_i}\nabla\cdot\bfq(\bfu) d\bfx\right|,
  \quad 
  %D_i = ||\bfv(\bfu_i)||_{\ell^2}\left|\left|\int_{S_i}\nabla\cdot\bff(\bfu)d\bfx\right|\right|_{\ell^2}
  %+\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|.
  D_i = \sum_{k=1}^{d+1}|\bfv_k(\bfu_i)|\left|\int_{S_i}\left(\nabla\cdot\bff(\bfu)\right)_kd\bfx\right|
  +\left|\int_{S_i}\nabla\cdot\bfq(\bfu)d\bfx\right|,
\end{align*}
where $({\bf z})_k$ denotes the $k$-th component of ${\bf z}\in\mathbb{R}^{d+1}$ and $d$ is the number of physical 
dimensions. In the next section, we use $R_i$ to blend Rusanov's and Roe's Riemann solvers.

\subsection{Blended Riemann solver}\label{sec:blended_rs}
In this section, we use \eqref{Ri}, which is a normalized measurement of the entropy residual \eqref{ent_residual}, 
to blend the Rusanov's and the Roe's Riemann solvers. To do this, we define the following numerical flux:
\begin{align}\label{num_flux_ev}
  F_{i+1/2} = \frac{\bff(\bfu_{i+1})+\bff(\bfu_i)}{2} 
  - \frac{1}{2} \left( R_{i+1/2}\lambda_{i+1/2}^{\max}\mathbb{I} + (1-R_{i+1/2})|\hat A_{i+1/2}| \right)(\bfu_{i+1}-\bfu_{i}),
\end{align}
and similarly for $F_{i-1/2}$. Here $\mathbb{I}\in\mathbb{R}^{3\times 3}$ is the identity matrix,
$R_{i+1/2}=\max\{R_{i+1},R_i\}$, $\lambda_{i+1/2}^{\max}$ is the 
upper bound on the wave speed used in the Rusanov's Riemann solver, see \S \ref{sec:rusanov}, 
and $\hat A_{i+1/2}$ is the Roe's averaged flux Jacobian, see \S \ref{sec:roe}.

%In terms of fluctuations, 
The first-order method is given simply by \eqref{first-order_FV} 
with the numerical flux $F_{i+1/2}$ given by \eqref{num_flux_ev}. 
In smooth regions $R_{i+1/2}\approx 0$ so the numerical flux is essentially the Roe's numerical flux. 
Near discontinuities or steep gradients, $R_{i+1/2}$ becomes large. If, in particular, $R_{i+1/2}\approx 1$, 
then the numerical flux is essentially the Rusanov's numerical flux.  
%
The first-order method can also be written in terms of fluctuations via \eqref{first-order_via_fluct}. 
To do this we consider the waves ($\W^p_{i\pm 1/2}$, with $p=1,2,3$) 
from the Roe's solver and define 
\begin{align}\label{lambda_p}
  \lambda_{i+1/2}^p := R_{i+1/2}\lambda_{i+1/2}^{\max} + (1-R_{i+1/2})|\hat \lambda_{i+1/2}^p|,
\end{align}
where $\hat\lambda_{i+1/2}^p$ is the $p$-th eigenvalue of the Roe's solver corresponding to the 
interface $i+1/2$. Note that $\lambda_{i+1/2}^p\geq 0$.
The fluctuations are then given by 
\begin{subequations}\label{ev_fluctuations}
\begin{align}
  \A^+\Delta Q_{i-1/2}&=\frac{1}{2}\sum_p \left(\hat\lambda_{i-1/2}^p + \lambda_{i-1/2}^p\right)\W_{i-1/2}^p, \\
  \A^-\Delta Q_{i+1/2}&=\frac{1}{2}\sum_p \left(\hat\lambda_{i+1/2}^p - \lambda_{i+1/2}^p\right)\W_{i+1/2}^p.
\end{align}
\end{subequations}

To implement the second-order Lax-Wendroff-LeVeque type method in Clawpack, the Riemann solver must provide 
the fluctuations, the waves and their speed. The fluctuations are simply \eqref{ev_fluctuations} 
and the waves are the Roe's waves $\W_{i+1/2}^p$, with $p=1,2,3$. 
In the framework of LeVeque, for a given interface $i+1/2$, 
we would need to consider a wave speed $s_{i+1/2}^p$ so that 
\begin{align*}
  \hat \lambda_{i+1/2}^p+\lambda_{i+1/2}^p = s_{i+1/2}^p+|s_{i+1/2}^p|, \qquad 
  \hat \lambda_{i+1/2}^p-\lambda_{i+1/2}^p = s_{i+1/2}^p-|s_{i+1/2}^p|,
\end{align*}
which is not possible to obtain in general (since the problem is overdetermined). 
Instead, we consider 
$s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$ 
 and study the consequences of this choice once the second-order corrections are applied.
Here $\sgn(z)$ is the sign function of $z$. 
The second-order Lax-Wendroff-LeVeque method for the one-dimensional problem is given by 
\begin{align}\label{second-order}
  u_i^{n+1}=u_i^n
  -\frac{\Delta t}{\Delta x}\left[\A^-\Delta Q_{i+1/2}+\A^+\Delta Q_{i-1/2}\right]
  -\frac{\Delta t}{\Delta x}\left[\tilde F_{i+1/2}-\tilde F_{i-1/2}\right],
\end{align}
where 
\begin{align*}
  \tilde F_{i+1/2} = \frac{1}{2}\sum_p\lambda_{i+1/2}^p
  \left(1-\frac{\Delta t}{\Delta x}\lambda_{i+1/2}^p\right)\tilde\W_{i+1/2}^p,
\end{align*}
and similarly for $\tilde F_{i-1/2}$, are the flux corrections with $\tilde\W_{i+1/2}^p$ 
being a TVD-limited version of $\W_{i+1/2}^p$.

To understand the effect of choosing $s_{i+1/2}^p=\sgn(\hat\lambda_{i+1/2}^p)\lambda_{i+1/2}^p$ 
as the speed of $\W_{i+1/2}^p$, 
let us assume that no limiters are applied; i.e., $\tilde\W_{i\pm 1/2}^p\equiv \W_{i\pm 1/2}^p$.
By doing this, equation \eqref{second-order} can be written as 
\begin{align*}
  u_i^{n+1}=u_i^n
  &-\frac{\Delta t}{2 \Delta x}
  \sum_p
  \left[\left(\hat\lambda_{i+1/2}^p -\frac{\Delta t}{\Delta x}(\lambda_{i+1/2}^p)^2 \right)\W_{i+1/2}^p
  +
  \left(\hat\lambda_{i-1/2}^p +\frac{\Delta t}{\Delta x}(\lambda_{i-1/2}^p)^2 \right)\W_{i-1/2}^p\right],
\end{align*}
which corresponds to the standard Lax-Wendroff method written in terms of fluctuation-like quantities. 
Note that if $R_{i\pm 1/2}=0$, we recover the standard second-order Lax-Wendroff method based on 
Roe's Riemann solver. 
In the numerical experiments in \S\ref{sec:num}, we obtain second-order experimental 
order of convergence (EOC) for problems with smooth solution, which is expected since 
$R_{i\pm 1/2}\approx 0$ in smooth regions.
Similarly, if $R_{i\pm 1/2}=1$, we recover the Lax-Wendroff method based on Rusanov's Riemann solver, 
which is clear by writting the fluctuations in Rusanov's method in terms of Roe's waves and speeds, 
see \eqref{fluct_rus}.
